\emph{icepack}: a novel glacier flow modeling package

### 2020-11-23 icepack.tex
Change use of diagnostic + prognostic model vs mass + momentum transport

\subsection{Mass transport}

The PDE describing how the ice thickness changes in time is a conservative advection equation:
\begin{equation}
    \frac{\partial h}{\partial t} + \nabla\cdot hu = \dot a_s - \dot a_b,
\end{equation}
where $\dot a_s$ and $\dot a_b$ are the surface and basal mass balance.
The thickness is represented using continuous, piecewise polynomial basis functions in each cell of the mesh.
The ice thickness and bed elevation are used to update the surface elevation at every timestep, and we need the surface gradient to calculate the driving stress.
The current mass transport implementation in icepack assumes continuous basis functions because this makes for the most straightforward calculation of the ice surface slope, which is part of the driving stress.
We have not yet implemented a solver that works for discontinuous basis functions, but this extension is completely feasible within our framework and may come in a future release of icepack.

@@

\subsection{Momentum transport}

There are four diagnostic models implemented in icepack.
See \citet{dukowicz2010consistent} for a thorough explanation of the various approximations and action principles velow.

@@

All the momentum transport models inherit a fundamental nonlinearity in the mechanics of ice flow.
For a Newtonian viscous fluid, the stress tensor $\tau$ and the strain rate tensor $\dot\varepsilon$ are linearly proportional to each other.
Glaciers, however, have a nonlinear constitutive relation:
\begin{equation}
    \dot\varepsilon = A|\tau|^{n - 1}\tau
\end{equation}
where $A$ is the \emph{fluidity} and $n$ is the \emph{Glen flow law exponent} \citep{cuffey2010physics}.
The most commonly-used value of the flow law exponent is $n \approx 3$ as determined from laboratory experiments.

@@

Every momentum transport model in icepack is encapsulated in its own Python class.
The key responsibility of the model classes is to take in the input fields -- the ice thickness, velocity, etc. -- and return a symbolic description of the action functional in UFL.

@@

The shallow stream model can be derived from the Blatter-Pattyn approximation by assuming nearly plug flow, i.e. $\partial u/\partial z \ll \partial u/\partial x$.
In this case, the momentum equation can be vertically integrated to obtain a 2D system of equations.
The class in icepack that represents this physics model is called \texttt{IceStream}.
The terms in the action functional are:
\begin{align}
    \text{viscosity} & = \frac{n}{n + 1}\int_\Omega hA^{-\frac{1}{n}}|\dot\varepsilon(u)|^{\frac{1}{n} + 1}\ud x \\
    \text{friction} & = \frac{m}{m + 1}\int_\Omega C|u|^{\frac{1}{m} + 1}\ud x \\
    \text{gravity} & = -\int_\Omega\rho_I gh\nabla s\cdot u\ud x \\
    \text{terminus} & = \frac{1}{2}\int_\Gamma(\rho_I gh^2 - \rho_Wgd^2)u\cdot \nu\ud\gamma
\end{align}
and the action functional is
\begin{equation}
    J = \text{viscosity} + \text{friction} - \text{gravity} - \text{terminus}.
\end{equation}
When the ice is floating, there are two simplifications: the friction coefficient $C$ is 0, and the surface elevation $s$ can be written in terms of the thickness $h$ as
\begin{equation}
    s = (1 - \rho_I / \rho_W)h.
\end{equation}
Additionally, the terminal stress term of the action disappears by applying integration by parts to the gravity term.
Since the action functional for ice shelves has fewer terms than for grounded ice streams, we have defined a separate \texttt{IceShelf} model class.
The ice shelf and ice stream models share common components, i.e. the viscosity and side wall stress.

@@

The simplest explicit timestepping schemes are unstable with continuous Galerkin finite elements.
To solve the mass transport equation, other packages use the streamlined upwind Petrov-Galerkin (SUPG) method for the timestepping scheme \citep{brinkerhoff2013data, larour2012continental}.
The SUPG method with an explicit time discretization is conditionally stable \citep{donea2003finite}.
This scheme introduces a tuneable stabilization parameter, so in the interest of simplicity we instead default to the implicit Euler scheme.
The implicit Euler scheme requires solving a non-symmetric linear system, but the computational cost comes with the advantage of unconditional stability.

Practicing glaciologists who have not studied numerical PDE may be unfamiliar with the Courant-Friedrichs-Lewy condition.
The use of an unconditionally stable scheme guarantees that they will get an answer should they try to use a large timestep rather than a runtime error.
The extra cost of using an implicit time discretization for the mass transport equation is dwarfed by the cost of the diagnostic solve in any case.
Advanced users who are interested in maximizing performance can subclass the mass transport solver to implement a faster explicit scheme.

>>>>

\subsection{Prognostic model}

The \emph{prognostic model} or \emph{mass transport equation} (the two terms are synonymous) describes how the ice thickness changes in time.
The prognostic model is a conservative advection equation:
\begin{equation}
    \frac{\partial h}{\partial t} + \nabla\cdot hu = \dot a_s - \dot a_b,
\end{equation}
where $\dot a_s$ and $\dot a_b$ are the surface and basal mass balance.
The thickness is represented using continuous, piecewise polynomial basis functions in each cell of the mesh.
The ice thickness and bed elevation are used to update the surface elevation at every timestep, and we need the surface gradient to calculate the driving stress.
The current prognostic model implementation in icepack assumes continuous basis functions because this makes for the most straightforward calculation of the ice surface slope, which is part of the driving stress.
We have not yet implemented a solver that works for discontinuous basis functions, but this extension is completely feasible within our framework and may come in a future release of icepack.

@@

\subsection{Diagnostic model}

There are four \emph{diagnostic} or \emph{momentum transport} models implemented in icepack.
See \citet{dukowicz2010consistent} for a thorough explanation of the various approximations and action principles velow.

@@

All the diagnostic models inherit a fundamental nonlinearity in the mechanics of ice flow.
For a Newtonian viscous fluid, the stress tensor $\tau$ and the strain rate tensor $\dot\varepsilon$ are linearly proportional to each other.
Glaciers, however, have a nonlinear constitutive relation:
\begin{equation}
    \dot\varepsilon = A|\tau|^{n - 1}\tau
\end{equation}
where $A$ is the \emph{fluidity} and $n$ is the \emph{Glen flow law exponent} \citep{cuffey2010physics}.
The most commonly-used value of the flow law exponent is $n \approx 3$ as determined from laboratory experiments.

@@

Every diagnostic model in icepack is encapsulated in its own Python class.
The key responsibility of the model classes is to take in the input fields -- the ice thickness, velocity, etc. -- and return a symbolic description of the action functional in UFL.

@@

The shallow stream model can be derived from the Blatter-Pattyn approximation by assuming nearly plug flow, i.e. $\partial u/\partial z \ll \partial u/\partial x$.
The momentum equations can, again, be vertically integrated to obtain a 2D system of equations.
The class in icepack that represents this physics model is called \texttt{IceStream}.
The terms in the action functional are:
\begin{align}
    \text{viscosity} & = \frac{n}{n + 1}\int_\Omega hA^{-\frac{1}{n}}|\dot\varepsilon(u)|^{\frac{1}{n} + 1}\ud x \\
    \text{friction} & = \frac{m}{m + 1}\int_\Omega C|u|^{\frac{1}{m} + 1}\ud x \\
    \text{gravity} & = -\int_\Omega\rho_I gh\nabla s\cdot u\ud x \\
    \text{terminus} & = \frac{1}{2}\int_\Gamma(\rho_I gh^2 - \rho_Wgd^2)u\cdot \nu\ud\gamma
\end{align}
and the action functional is
\begin{equation}
    J = \text{viscosity} + \text{friction} - \text{gravity} - \text{terminus}.
\end{equation}
When the ice is floating, there are two simplifications: the friction coefficient $C$ is 0, and the surface elevation $s$ can be written in terms of the thickness $h$ as
\begin{equation}
    s = (1 - \rho_I / \rho_W)h.
\end{equation}
Additionally, the terminal stress term of the action disappears by applying integration by parts to the gravity term.
Since the action functional for ice shelves has fewer terms than for grounded ice streams, we have defined a separate \texttt{IceShelf} model class.
The ice shelf and ice stream models share common components, i.e. the viscosity and side wall stress.

@@

The simplest explicit timestepping schemes are unstable with continuous Galerkin finite elements.
To solve the advection equation equation, other packages use the streamlined upwind Petrov-Galerkin (SUPG) method for the timestepping scheme \citep{brinkerhoff2013data, larour2012continental}.
The SUPG method with an explicit time discretization is conditionally stable \citep{donea2003finite}.
This scheme introduces a tuneable stabilization parameter, so in the interest of simplicity we instead default to the implicit Euler scheme.
The implicit Euler scheme requires solving a non-symmetric linear system, but the computational cost comes with the advantage of unconditional stability.

Practicing glaciologists who have not studied numerical PDE may be unfamiliar with the Courant-Friedrichs-Lewy condition.
The use of an unconditionally stable scheme guarantees that they will get an answer should they try to use a large timestep rather than a runtime error.
The extra cost of using an implicit time discretization for the prognostic equation is dwarfed by the cost of the diagnostic solve in any case.
Advanced users who are interested in maximizing performance can subclass the solver to implement a faster explicit scheme.
